| Early stopping in clinical PET studies: \n| how to reduce expense and exposure

### 2020-10-12 ArticleDraft/Draft.Rmd
Updated draft after pre-print feedback

---
title: "| Early stopping in clinical PET studies: \n| how to reduce expense and exposure"
author: "| Jonas Svensson^1^, Martin Schain^2^, Gitte M. Knudsen^2,3^, Todd Ogden^4,5^, Pontus Plavén-Sigray^2^*"
date: ' '
output:
  pdf_document: default
  word_document: default
  html_document: default
header-includes:
    - \usepackage[most]{tcolorbox}
    - \usepackage{setspace}\doublespacing
    - \usepackage[hang,flushmargin]{footmisc} 
    - \usepackage{units}
csl: /Users/pontus.sigray/Github/Early_stopping_in_PET/ArticleDraft/csl/ejnmmi-research.csl
bibliography: /Users/pontus.sigray/Github/Early_stopping_in_PET/ArticleDraft/BibTex/Early_stopping_in_clinical_PET_research.bib
---

@@

One way to avoid superfluous PET scans is to intermittently check for a statistical effect while the study is still ongoing, generally termed *sequential testing* of data. In the classical NHST framework, however, sequential testing does pose a problem, as it can greatly inflate the nominal type I error rate. If a new uncorrected p-value is calculated and used for making inference after the collection of each patient-control pair, the type I error rate will be above 20% for commonly seen sample sizes in PET literature [@Strube2006; @Albers2019].

@@

There is however a different metric that can be used to test scientific hypotheses, and that has been gaining traction in the field of biomedicine during the last decade: the Bayes Factor (BF) [@Jeffreys1961; @Kass1995; @Lee2014; @Schonbrodt2018]. The BF has two advantages compared to the classical NHST procedure: 1) it is well suited for sequential testing of data [@Rouder2014; @Schonbrodt2017] and 2) it allows for quantification of relative evidence in favour of H~0~, meaning that a PET study can be stopped when it is determined that H~0~ is supported [@Dienes2014]. In this tutorial we will show how to use sequential BF tests in common PET study designs, in order to stop data collection in a study early.

@@

In contrast to the classical NHST procedure, sequential testing and optional stopping does not pose the same challenge for Bayesian hypothesis testing framework [@Rouder2014; @Schonbrodt2017]. In sequential BF testing, the null hypothesis is assessed in constant competition to the alternative hypothesis. If a BF is calculated after the collection of each data point, it informs the researcher how the stated hypotheses are gaining or losing support from the data. Because of this, the BF can be used for “online” monitoring of incoming data. The researchers can stop when they reach a pre-set decision threshold, a pre-set maximum sample size (or when they simply run out of money, time or patience) [@Rouder2014; @Schonbrodt2017].

@@

In this tutorial we will explore how to use the default BF t-test to sequentially test data in two common clinical PET study designs: a cross-sectional (e.g., a patient-control comparison) and a paired (e.g., a pre-intervention-post scan) design. Our first aim is to assess whether sequential BF testing can be used to stop data collection at an earlier stage compared to the classical NHST framework, while still keeping the number of false positives below the commonly set threshold of 5%. Our second aim is to assess whether sequential BF testing can be used to stop early, not only when there is an effect in the population (H~1~ is true), but also when there is no effect in the population (H~0~ is true). Our third aim is to apply sequential BF testing on real clinical PET data, and evaluate the outcome.

@@

---
# References for Table 2 below: 
---

@@

knitr::kable(x = tab2.df,
             format = 'latex', 
             booktabs = T, 
             align = 'c', 
             escape = F,
             col.names = linebreak(c('Effect size',
                           '$[^{11}$C]raclopride \n BP\\textsubscript{ND} (\\%)  difference',
                           '$[^{11}$C]AZ10419369 \n BP\\textsubscript{ND} (\\%) difference',
                           '$[^{11}$C]PBR28 \n V\\textsubscript{T} (\\%) difference',
                           '$[^{11}$C]UCB-J \n V\\textsubscript{T} (\\%) difference'),align = 'c'),
      caption = 'The standardized effect size used to assess the difference between two groups (Cohen’s D for a cross-sectional design) and difference within the same subjects (Cohen’s Dz for a paired design) translated into raw difference in BP\\textsubscript{ND} and V\\textsubscript{T} and percentage difference (\\%) for four PET radioligands. The mean, SD between subjects and SD within subjects for each radioligand were taken from test-retest studies on healthy subjects. NB: the variance is likely to be higher in a more heterogeneous clinical population, which will lead to smaller effect sizes for the same raw or \\% difference. ') %>%
  kable_styling(latex_options = c("hold_position"),
                full_width = F,
                position = "center") %>%
     kableExtra::group_rows(group_label = "Cohen's D",1,5) %>%
     kableExtra::group_rows(group_label = "Cohen's Dz",6,10)

@@

Our first aim of this tutorial is to compare sequential BF testing to the conventional way of evaluating research hypotheses in a clinical PET study, which is to apply a classical NHST test after a fixed sample size has been collected. The objective is to show how sequential BF testing can be used in order to stop a study early while still keeping the rate of false positive findings under the commonly set threshold of 5%. In order to compare the sequential BF framework to classical NHST with a fixed sample size, we will only stop data collection when there is evidence in favour of an effect. In this simulation, evidence for the null hypothesis will hence be treated the same as inconclusive results, i.e. no early stopping decisions will be made when BF < 1.

@@

Figure \ref{fig_k_at_5fpr}A and 3B show the estimated false positive rate when applying sequential BF testing in the case in which there is no difference between the two groups. Three different stopping thresholds have been used (BF > 3, 4 and 6, respectively). When the maximum allowed sample size increases, the false positive rate goes up. This is because the longer the sequential testing can go on, the more decisions are being made. Some of these decisions will be wrong, meaning that the researcher will claim support for an effect, when in fact there is no effect in the population. It is not possible to keep false positives below the 5% using a decision threshold of 4 with a maximum sample above 30 subjects/group. Instead, if the researchers wish to control the false positives at the 5% level, one option is to increase the decision threshold. Figure \ref{fig_k_at_5fpr}C and 3D show the estimated decision threshold that would be needed to keep the false positive rate at 5% when using sequential testing with the default BF t-test, while still starting at 12 subjects/group and checking the BF after the collection of each patient-control pair.

\begin{figure}
        \centering
        \includegraphics[width=140mm]{/Users/pontus.sigray/Github/Early_stopping_in_PET/Results/Figures/Figure4.png}
        \caption{\textit{Panels A and B show the proportion of studies that showed support for H\textsubscript{1} (aka “power curves”) for BF sequential testing (blue) and fixed N NHST (black), at different population effects (starting at D = $0$, i.e. no effect). For NHST, only one test is performed at N = $30$ subjects/group. For the sequential testing, $12$ subjects/group are first collected, then BF is checked after each added comparison pair until $30$ subjects/group is reached, using a stopping threshold of $4$. Panels C and D show the average number of subjects needed to reach a stopping decision at different population effects. NHST is the black line (fixed at N = $30$ subjects/group); BF sequential testing is the blue line with shaded area denoting $\pm$ $1$SD.}}
        \label{fig_powerPhase1}
\end{figure}

Figure \ref{fig_powerPhase1} shows the results from the simulations when Nmax=30 subjects/group, using a decision threshold of 4. In the upper panel the rate of positive studies at different population effects is visualised. The curves cross the y-axis at D=0 (i.e., the false positive rate), 5% for both BF seq and NHST. As can be seen, had the researcher instead applied a classical NHST t-test at the fixed sample size of 30 subjects/group, they would have increased the power to detect the population effect by no more than ~10% for any effect size compared to using sequential BF testing.

Figures 4C and 4D show the average number of subjects needed to reach a decision at different population effects. As the underlying difference between the two populations is increased, fewer subjects are needed to reach a decision using sequential BF testing. Already at a true population difference of Cohen’s Dz = 0.5, researchers will on average save 30% in terms of both expense and amount of injected radioactivity when applying sequential BF testing compared to the fixed N NHST procedure.

@@

The area around the point where the red and black line cross[^fn6], is a weak spot for the BF sequential testing procedure. At this range of true effects in the population, the risk of obtaining an inconclusive result is at its highest (i.e., the blue line is at its nadir). In the cases in which a decision is reached, the study will stop with equal probability for H~1~ and H~0~, and the risk for false negative evidence is around 25%.

Assuming that the PET researcher would be interested in stopping a study early when H~0~ is supported, a range of population effects with high risk of false negative evidence will always exist. It is therefore sensible to beforehand decide on a minimal population effect of interest, and choose the settings to ensure that any larger effect does not have too high a risk of generating evidence for H~0~. For example, with the settings presented in Figure \ref{fig_powerPhase2} (Nstart = 12, Nmax = 30, threshold = 4) it could be said that all effects between D = 0 and 0.45 ought to be of little clinical interest, in order for sequential BF testing to be applicable in this case since the risk of false negative evidence for these effects are is > 10%[^fn7].

@@

\begin{tcolorbox}[title=\textbf{Practical recommendations 2},title filled]
{In this simulation set-up we show that it is possible to stop a PET study early either when there is a true underlying effect in the population, or when the effect is zero. For a patient-control comparison, we recommend that sequential BF testing not start until at least 12 subjects per group have been collected, setting the BF threshold to a minimum of 3, and using a one-sided Cauchy distribution (centered around zero with a \textit{r} = 0.707) as the alternative hypothesis. Using a one-sided test means that the researcher must make an a-priori prediction of which direction the effect will have, and not change this prediction after the study is started, similar to performing a one-sided classical NHST t-test.}
\end{tcolorbox}

## Objective 3 - Application to real clinical PET data 
The third and final aim of this tutorial is to apply sequential BF testning to a real clinical PET setting. To this purpose, we used already collected data of patients with major depressive disorder and healthy control subjects examined with [11C]WAY100635, a radioligand, which binds to the serotonin 1A (5HT1A) receptor [@Parsey2010; @Chen2019]. From this data, we included 40 medication-free patients (mean Age 36.2 (12.9SD); 25 females) and 40 healthy controls (mean Age 37.1 (14.0SD), 25 females), using BP~F~ (specific binding over the free-fraction in plasma) as the primary outcome measure for assessing 5HT1A receptor availability in the brain. The 5HT1A receptor acts inhibitory on serotonergic neurons in the raphe nuclei. A high concentration of receptors in raphe will likely lead to lower transmitter release in serotonergic projection areas. Hence, we hypothesized that patients suffering from MDD show higher 5HT1A receptor availability in the raphe nuclei compared to healthy controls. Here we apply sequential BF testing to examine whether this hypothesis gains enough support to stop data collection before 40 subjects per group are reached.

###Methods
We applied sequential BF testing to assess the support in data in favor of patients having higher BP~F~ in the raphe nuclei than controls (H~1~), as compared to no difference (H~0~). We also examined the evidence in data in favor of H~0~ over H~1~, to stop the study early if the null hypothesis was supported. The stopping threshold was set to BF>5 for H~1~ (BF < $\nicefrac{1}{5}$ for H~0~), in order to keep the rate of false positive evidence below 5%[^fn8] (see Supplementary Figure S1). We used a one-sided Cauchy distribution, centered around 0 with an *r* = 0.707 to describe the predicted mean difference under H~1~. H~0~ was specified as the point zero value.

@@

When applying sequential BF testing, and the true population effect is small, the relative savings are modest compared to using the classical NHST procedure with fixed sample size. As the population effect becomes bigger however, it is possible to (on average) stop the data collection early, potentially saving more than half of the resources that would be needed for a PET study carried out with the conventional NHST approach. For a range of effects, NHST will however show around 5-10% higher true positive rate compared to sequential BF testing (Figure \ref{fig_powerPhase1}A-B). By using a lower BF decision threshold, the proportion of true positive stopping decisions will increase when there is an effect, but it comes with a cost of more false positives when H~0~ is true.

There are sequential testing methods other than BF that can be used to stop the data collection early, and offer exact control over error rates. For the NHST framework, a viable alternative is to apply a so-called group-sequential design [@Schulz2005; @Lakens2014]. In this tutorial, we did not evaluate such a procedure, but when used to stop early for an effect, it produces similar results to sequential BF testing [@Schonbrodt2017].

One advantage of sequential BF testing is that a PET study can also be stopped early when there is no effect in the population, i.e., when H~0~ is true. If the researcher wants to use a sequential BF to assess support for both H~1~ and H~0~, we recommend using a one-tailed default BF t-test. This entails that the researcher must make an a priori prediction on the direction of the effect (e.g., patients have higher binding than control subjects, or vice versa). Without using one-sided default BF t-test, it will be difficult (or even impossible) to stop a study early in support for H~0~, given the sample sizes used in typical PET studies (N<100 subjects/group) while still using a threshold that is high enough to keep the rate of false evidence at reasonable levels (see Supplementary Figure S11).

@@

Figure \ref{fig_powerPhase2} presents the true and false evidence rates the researchers can expect when using a one-sided BF t-test and when they are correct in their prediction of the direction of an effect. If there is a sizable patient-control difference in the population, but it goes in the opposite direction to the predicted one, then using this approach would lead studies to be stopped early due to the null-hypothesis being supported almost 100% of the time. Of course, the null is not true in this case, but it is closer to the true population effect than the a-priori defined alternative hypothesis. Hence, it will often gain the most support from the data.

When allowing stopping for H~0~ as well, the researcher should be aware that the false negative evidence rate can become large when the effect is close to zero, but not exactly zero. Using sequential BF testing, researchers will run the risk of stopping the study, claiming support for no effect, when there is in fact a true but small effect in the population. Whether or not this is acceptable to the researchers depends on how small an effect has to be, in order for the researcher to consider it practically indistinguishable from zero. Using the sequential BF approach with the settings described in Figure \ref{fig_powerPhase1} (e.g., a decision threshold of 4), the false negative evidence rate becomes high (>10%) for all effects between 0 and 0.45 for a cross-sectional design. Hence, if PET researchers consider this range of effects to be practically or clinically relevant, caution is warranted before applying a sequential BF testing procedure as shown here. Instead the researchers could specify H~1~ to assign larger plausibility to smaller effects, by lowering the width of the Cauchy (setting *r* to, e.g., 0.5) when planning the study. This will increase the rate of true positive evidence at smaller effect sizes, but it will also increase the rate of false positive evidence when H~0~ is true.

@@

\begin{tcolorbox}[title=\textbf{Example 3},title filled]
{There is a true difference in [$^{11}$C]raclopride BP\textsubscript{ND} between a patient and a control population that is of size Cohen’s D = 0.2 (a difference of 0.07 BP\textsubscript{ND} or 2\%, Table 2). For such a small effect, studies will often be stopped because H\textsubscript{0} is supported when applying the sequential BF testing procedure, e.g., when using a threshold of 4 and $\nicefrac{1}{4}$, respectivly. A PET researcher might consider such an effect to be practically indistinguishable from zero. In such a scenario, support for H\textsubscript{0} is hence of little concern, or even a prefered outcome for the researcher. On the other hand, if the researcher considers a Cohen’s D = 0.2 to be clinically relevant, then a sequential BF procedure using the default BF t-test cannot reliably give the correct stopping decision for H\textsubscript{1} in commonly seen sample sizes in PET research (<100 subjects/group). If the researcher wants to be able to detect such a small effect, they need to use a different specification of H\textsubscript{1}, or a higher decision threshold (e.g. BF>10) together with a much larger Nmax.}
\end{tcolorbox}

For any given PET-study where sequential BF testing is applied there will be a chance of reaching Nmax without crossing the pre-set decision threshold. In such cases, where the researcher ends up with an “inconclusive” finding, the BF is still interpretable. The suggested evidence thresholds in Table 1 can be used to report the support in data in favor of the two competing hypotheses, even though BF never formally reached the a priori set stopping threshold. For example, if the BF is above 3 (or below $\nicefrac{1}{3}$), this can still be reported as “moderate” support in data for one hypothesis over the other. However, in doing so the researcher cannot any longer say that they are controlling the rate of false evidence at a prespecified level.

@@

In addition to the settings presented in this tutorial, there are several modifications that can be made for the BF test. For example, different decision thresholds can be applied for H~1~ and H~0~ depending on whether the researcher thinks that false positives are more important to avoid than false negatives, or vice versa . If the researcher has an informed idea on what the effect should be, based on previous literature or bounds of a minimal effect size of interest, it is possible to use an informed alternative hypothesis instead of the default Cauchy distribution.

@@

It should by now be clear to the reader that a set of critical decisions needs to be made before applying sequential BF testing in a clinical PET study. For this reason, we recommend all researchers to pre-register their analysis before data collection starts [^fn9] [@VantVeer2016; @Poldrack2017; @Knudsen2020]. A pre-registration can be thought of as a safety net for the researcher. It helps guide the analysis and interpretation of data so that error rates are kept under control. It can also be shown to reviewers or readers to increase the credibility of the methods and findings. Deviations from a pre-registration is of course possible, and often warranted, but should be reported transparently in the article.

@@

If the reader is interested in learning more about sequential testing the following articles are a good start for a BF approach: @Schonbrodt2018, @Rouder2014 and @Schonbrodt2017; and for a NHST approach: @Schulz2005 and @Lakens2014.

>>>>

---
title: "| Early stopping in clinical PET studies: \n| how to reduce expense and exposure"
author: "| Jonas Svensson^1^, Martin Schain^2^, Gitte M. Knudsen^2,3^, Todd Ogden^4,5^, Pontus Plavén-Sigray^2^*"
date: ' '
output:
  pdf_document: default
  word_document: default
  html_document: default
always_allow_html: true
header-includes:
    - \usepackage[most]{tcolorbox}
    - \usepackage{setspace}\doublespacing
    - \usepackage[hang,flushmargin]{footmisc} 
    - \usepackage{units}
    - \usepackage{xcolor}
csl: /Users/pontus.sigray/Github/Early_stopping_in_PET/ArticleDraft/csl/ejnmmi-research.csl
bibliography: /Users/pontus.sigray/Github/Early_stopping_in_PET/ArticleDraft/BibTex/Early_stopping_in_clinical_PET_research.bib
---

@@

One way to avoid superfluous PET scans is to intermittently check for a statistical effect while the study is still ongoing, generally termed *sequential testing* of data. In the uncorrected NHST framework, however, sequential testing does pose a problem, as it can greatly inflate the nominal type I error rate. If a new uncorrected p-value is calculated and used for making inference after the collection of each patient-control pair, the type I error rate will be above 20% for commonly seen sample sizes in PET literature [@Strube2006; @Albers2019].

@@

There are however techniques that are appropriate to use when performing sequential testing. One possibility involves correcting the significance threshold so that the overall error rate does not surpass the alpha level when performing intermittent NHST tests [@Proschan2006] (see the Discussion for a more elaborate description of this procedure). An alternative possibility is to use an entirely different metric to test scientific hypotheses: the Bayes Factor (BF) [@Jeffreys1961; @Kass1995; @Lee2014; @Schonbrodt2018]. The BF has been gaining traction in  the field of biomedicine during the last decade and has two important characteristics; first, it is well suited for sequential testing of data [@Rouder2014; @Schonbrodt2017], and second, it also allows for quantification of relative evidence in favour of H~0~, meaning that a PET study can be stopped when it is determined that H~0~ is supported [@Dienes2014]. In this tutorial we will show how to use sequential BF tests in common PET study designs, in order to stop data collection in a study early.

@@

In sequential BF testing, the null hypothesis is assessed in constant competition to the alternative hypothesis. If a BF is calculated after the collection of each data point, it informs the researcher how the stated hypotheses are gaining or losing support from the data. Because of this, the BF can be used for “online” monitoring of incoming data. The researchers can stop when they reach a pre-set decision threshold, a pre-set maximum sample size (or when they simply run out of money, time or patience) [@Rouder2014; @Schonbrodt2017].

@@

In this tutorial we will explore how to use the default BF t-test to sequentially test data in two common clinical PET study designs: a cross-sectional (e.g., a patient-control comparison) and a paired (e.g., a pre-intervention-post scan) design. Our first aim is to demonstrate the relative merit of using sequential BF testing to stop data collection at an earlier stage, compared to simply applying a conventional "fixed N" design, while still keeping the number of false positives below the commonly set threshold of 5%. Our second aim is to assess whether sequential BF testing can be used to stop early, not only when there is an effect in the population (H~1~ is true), but also when there is no effect in the population (H~0~ is true). Our third aim is to apply sequential BF testing on real clinical PET data, and evaluate the outcome.

@@

@@

knitr::kable(x = tab2.df,
             format = 'latex', 
             booktabs = T, 
             align = 'c', 
             escape = F,
             col.names = linebreak(c('Effect size',
                           '$[^{11}$C]raclopride \n BP\\textsubscript{ND} (\\%)  difference',
                           '$[^{11}$C]AZ10419369 \n BP\\textsubscript{ND} (\\%) difference',
                           '$[^{11}$C]PBR28 \n V\\textsubscript{T} (\\%) difference',
                           '$[^{11}$C]UCB-J \n V\\textsubscript{T} (\\%) difference'),align = 'c'),
      caption = 'The standardized effect size used to assess the difference between two groups (Cohen’s D for a cross-sectional design) and difference within the same subjects (Cohen’s Dz for a paired design) translated into raw mean difference in BP\\textsubscript{ND} and V\\textsubscript{T} and percentage mean difference (\\%) for four PET radioligands. The mean, SD between subjects and SD within subjects for each radioligand were taken from test-retest studies on healthy subjects. NB: the variance is likely to be higher in a more heterogeneous clinical population, which will lead to smaller effect sizes for the same raw or \\% difference. ') %>%
  kable_styling(latex_options = c("hold_position"),
                full_width = F,
                position = "center") %>%
     kableExtra::group_rows(group_label = "Cohen's D",1,5) %>%
     kableExtra::group_rows(group_label = "Cohen's Dz",6,10)

@@

Our first aim of this tutorial is to compare sequential BF testing to the conventional way of evaluating research hypotheses in a clinical PET study, which is to apply a single test after the entire sample size has been collected (i.e., a "fixed N approach") [^fn6]. Here we will explore the differences between sequential BF testing and taking a fixed N approach. The aim is to present the reader with the trade-offs that are made when using a sequential procedure. We will therefore explore how sequential BF testing can be used in order to stop a study early while still keeping the rate of false positive findings under the commonly set threshold of 5%. In order to compare the sequential BF framework to the conventional approach, we will only stop data collection when there is evidence in favour of an effect. In this simulation, evidence for the null hypothesis will hence be treated the same as inconclusive results, i.e. no early stopping decisions will be made when BF < 1. For a demonstration on how to stop a study early also when H~0~ is supported, see Objective 2 below.

@@

Figure \ref{fig_k_at_5fpr}A and 3B show the estimated false positive rate when applying sequential BF testing in the case in which there is no difference between the two groups. Three different stopping thresholds have been used (BF > 3, 4 and 6, respectively). When the maximum allowed sample size increases, the false positive rate goes up. This is because the longer the sequential testing can go on, the more decisions are being made. Some of these decisions will be wrong, meaning that the researcher will claim support for an effect, when in fact there is no effect in the population. Eventuelly, when N becomes high enough, the rate of false positives will reach a reach an asymptote (see Supplementary Figure S1). But with a maximum sample size above 30 subjects/group, it is not possible to keep false positives below the 5% using a decision threshold of 4. Instead, if the researchers wish to control the false positives at the 5% level, one option is to increase the decision threshold. Figure \ref{fig_k_at_5fpr}C and 3D show the estimated decision threshold that would be needed to keep the false positive rate at 5% when using sequential testing with the default BF t-test, while still starting at 12 subjects/group and checking the BF after the collection of each patient-control pair.

\begin{figure}
        \centering
        \includegraphics[width=140mm]{/Users/pontus.sigray/Github/Early_stopping_in_PET/Results/Figures/Figure4.png}
        \caption{\textit{Panels A and B show the proportion of studies that showed support for H\textsubscript{1} (aka “power curves”) for BF sequential testing (blue) and fixed N test (black), at different population effects (starting at D = $0$, i.e. no effect). For the fixed N approach, only one test is performed at N = $30$ subjects/group. For the sequential testing, $12$ subjects/group are first collected, then BF is checked after each added comparison pair until $30$ subjects/group is reached, using a stopping threshold of $4$. Panels C and D show the average number of subjects needed to reach a stopping decision at different population effects. The fixed N test is the black line (N = $30$ subjects/group); BF sequential testing is the blue line with shaded area denoting $\pm$ $1$SD.}}
        \label{fig_powerPhase1}
\end{figure}

Figure \ref{fig_powerPhase1} shows the results from the simulations when Nmax=30 subjects/group, using a decision threshold just above 4. In the upper panel the rate of positive studies at different population effects is visualised. The curves cross the y-axis at D=0 (i.e., the false positive rate), 5% for both BF seq and the conventional fixed N test. As can be seen, had the researcher instead applied a fixed N approach (at 30 subjects/group), they would have increased the power to detect the population effect by no more than ~10% for any effect size compared to using sequential testing. Note that when calibrating the BF threshold to correspond to a 5% false positive rate, and stopping only for H~1~, the BF procedure will show identical results to performing sequential NHST t-tests where the significance level has been corrected (see Supplementary Figure SXXX).

Figures 4C and 4D show the average number of subjects needed to reach a decision at different population effects. As the underlying difference between the two populations is increased, fewer subjects are needed to reach a decision using sequential BF testing. Already at a true population difference of Cohen’s Dz = 0.5, researchers will on average save 30% in terms of both expense and amount of injected radioactivity when applying sequential BF testing compared to the fixed N approach.

In summary, if the PET researchers apply a sequential testing procedure as presented here, they have the potential of reducing both expense and radioactivity exposure, but it comes at a cost of a lower rate of true positives (i.e., statistical power), for a range of effect sizes.

@@

The area around the point where the red and black line cross[^fn7], is a weak spot for the BF sequential testing procedure. At this range of true effects in the population, the risk of obtaining an inconclusive result is at its highest (i.e., the blue line is at its nadir). In the cases in which a decision is reached, the study will stop with equal probability for H~1~ and H~0~, and the risk for false negative evidence is around 25%.

Assuming that the PET researcher would be interested in stopping a study early when H~0~ is supported, a range of population effects with high risk of false negative evidence will always exist. It is therefore sensible to beforehand decide on a minimal population effect of interest, and choose the settings to ensure that any larger effect does not have too high a risk of generating evidence for H~0~. For example, with the settings presented in Figure \ref{fig_powerPhase2} (Nstart = 12, Nmax = 30, threshold = 4) it could be said that all effects between D = 0 and 0.45 ought to be of little clinical interest, in order for sequential BF testing to be applicable in this case since the risk of false negative evidence for these effects is > 10%[^fn8].

@@

\begin{tcolorbox}[title=\textbf{Practical recommendations 2},title filled]
{In this simulation set-up we show that it is possible to stop a PET study early either when there is a true underlying effect in the population, or when the effect is zero. For a patient-control comparison, we recommend that sequential BF testing not start until at least 12 subjects per group have been collected, setting the BF threshold to a minimum of 3, and using a one-sided Cauchy distribution (centered around zero with a \textit{r} = 0.707) as the alternative hypothesis. Using a one-sided test means that the researcher must make an a priori prediction of which direction the effect will have, and not change this prediction after the study is started, similar to performing a classical one-sided NHST t-test.}
\end{tcolorbox}

## Objective 3 - Application to real clinical PET data 
The third and final aim of this tutorial is to apply sequential BF testing to a real clinical PET setting. To this purpose, we used already collected data of patients with major depressive disorder and healthy control subjects examined with [11C]WAY100635, a radioligand, which binds to the serotonin 1A (5HT1A) receptor [@Parsey2010; @Chen2019]. From this data, we included 40 medication-free patients (mean Age 36.2 (12.9SD); 25 females) and 40 healthy controls (mean Age 37.1 (14.0SD), 25 females), using BP~F~ (specific binding over the free-fraction in plasma) as the primary outcome measure for assessing 5HT1A receptor availability in the brain. The 5HT1A receptor acts inhibitory on serotonergic neurons in the raphe nuclei. A high concentration of receptors in raphe will likely lead to lower transmitter release in serotonergic projection areas. Hence, we hypothesized that patients suffering from MDD show higher 5HT1A receptor availability in the raphe nuclei compared to healthy controls. Here we apply sequential BF testing to examine whether this hypothesis gains enough support to stop data collection before 40 subjects per group are reached.

###Methods
We applied sequential BF testing to assess the support in data in favor of patients having higher BP~F~ in the raphe nuclei than controls (H~1~), as compared to no difference (H~0~). We also examined the evidence in data in favor of H~0~ over H~1~, to stop the study early if the null hypothesis was supported. The stopping threshold was set to BF>5 for H~1~ (BF < $\nicefrac{1}{5}$ for H~0~), in order to keep the rate of false positive evidence below 5%[^fn9] (see Supplementary Figure S1). We used a one-sided Cauchy distribution, centered around 0 with an *r* = 0.707 to describe the predicted mean difference under H~1~. H~0~ was specified as the point zero value.

@@

When applying sequential BF testing, and the true population effect is small, the relative savings are modest compared to using the conventional design where only one test is performed at a fixed sample size. As the population effect becomes bigger however, it is possible to (on average) stop the data collection early, potentially saving more than half of the resources that would be needed for a PET study carried out with the conventional approach. For a range of effects, a fixed N test will however show around 5-10% higher true positive rate compared to sequential BF testing (Figure \ref{fig_powerPhase1}A-B). By using a lower BF decision threshold, the proportion of true positive stopping decisions will increase when there is an effect, but it comes with a cost of more false positives when H~0~ is true.

#### Alternative sequential procedures
There are sequential testing methods other than BF that can be used to stop the data collection early, and that offer exact control over error rates. For the NHST framework, a viable alternative is to apply a so-called group-sequential NHST design [@Schulz2005]. In fact, when calibrating the BF threshold (using a "default" Cauchy-distribution for describing H~1~) to a 5% false positive rate when stopping only for H~1~ (see  Objective 1), the sequential BF procedure will be identical to performing corrected sequential NHST t-tests (supplementary Figure XXX). For NHST, there are several different efficient strategies for doing sequential testing to stop a study as soon as H~0~ can be rejected [@Proschan2006; @Lakens2014]. One alternative is to use "alpha-spending functions", where the significance threshold changes for each sequential test, making it harder to reject H~0~ at an early stage of the study, but easier later on (or vice versa), while still keeping the total false positive rate below 5% [@Demets1994; @Selwyn2004]. There are also methods for performing so called "futility testing", with which the researchers can stop a study early if it is deemed to  unlikely that a significant finding will be achieved later on [@Lachin2005]. Addressing such sequential-NHST procedures for application in clinical PET research are however outside the scope of this tutorial.

#### Evidence in favor of H~0~
One advantage of sequential BF testing is that a PET study can also be stopped early when there is no effect in the population, i.e., when H~0~ is true. If the researcher wants to use a sequential BF to assess support for both H~1~ and H~0~, we recommend using a one-tailed default BF t-test. This entails that the researcher must make an a priori prediction on the direction of the effect (e.g., patients have higher binding than control subjects, or vice versa). Without using one-sided default BF t-test, it will be difficult (or even impossible) to stop a study early in support for H~0~, given the sample sizes used in typical PET studies (N<100 subjects/group) while still using a threshold that is high enough to keep the rate of false evidence at reasonable levels (see Supplementary Figure S11).

@@

Figure \ref{fig_powerPhase2} presents the true and false evidence rates the researchers can expect when using a one-sided BF t-test and when they are correct in their prediction of the direction of an effect. If there is a sizable patient-control difference in the population, but it goes in the opposite direction to the predicted one, then using this approach would lead studies to be stopped early due to the null-hypothesis being supported almost 100% of the time. Of course, the null is not true in this case, but it is closer to the true population effect than the a priori defined alternative hypothesis. Hence, it will often gain the most support from the data.

When allowing stopping for H~0~ as well, the researcher should be aware that the false negative evidence rate can become large when the effect is close to zero, but not exactly zero. Using sequential BF testing, researchers will run the risk of stopping the study, claiming support for no effect, when there is in fact a true but small effect in the population. Whether or not this is acceptable to the researchers depends on how small an effect has to be, in order for the researcher to consider it practically indistinguishable from zero. Using the sequential BF approach with the settings described in Figure \ref{fig_powerPhase1} (e.g., a decision threshold of 4), the false negative evidence rate becomes high (>10%) for all effects between 0 and 0.45 for a cross-sectional design. Hence, if PET researchers consider this range of effects to be practically or clinically relevant, caution is warranted before applying a sequential BF testing procedure as shown here. Instead the researchers could increase the decision threshold, or specify H~1~ to assign larger plausibility to smaller effects, by lowering the width of the Cauchy (setting *r* to, e.g., 0.5) when planning the study. The latter will increase the rate of true positive evidence at smaller effect sizes, but it will also increase the rate of false positive evidence when H~0~ is true.

@@

\begin{tcolorbox}[title=\textbf{Example 3},title filled]
{There is a true difference in [$^{11}$C]raclopride BP\textsubscript{ND} between a patient and a control population that is of size Cohen’s D = 0.2 (a difference of 0.07 BP\textsubscript{ND} or 2\%, Table 2). For such a small effect, studies will often be stopped because H\textsubscript{0} is supported when applying the sequential BF testing procedure, e.g., when using a threshold of 4 and $\nicefrac{1}{4}$, respectively. A PET researcher might consider such an effect to be practically indistinguishable from zero. In such a scenario, support for H\textsubscript{0} is hence of little concern, or even a preferred outcome for the researcher. On the other hand, if the researcher considers a Cohen’s D = 0.2 to be clinically relevant, then a sequential BF procedure using the default BF t-test cannot reliably give the correct stopping decision for H\textsubscript{1} in commonly seen sample sizes in PET research (<100 subjects/group). If the researcher wants to be able to detect such a small effect, they need to use a different specification of H\textsubscript{1}, or a higher decision threshold (e.g. BF>10) together with a much larger Nmax.}
\end{tcolorbox}

#### Inconclusive findings 
For any given PET-study where sequential BF testing is applied there will be a chance of reaching Nmax without crossing the pre-set decision threshold. In such cases, where the researcher ends up with an “inconclusive” finding, the BF is still interpretable. The suggested evidence thresholds in Table 1 can be used to report the support in data in favor of the two competing hypotheses, even though BF never formally reached the a priori set stopping threshold. For example, if the BF is above 3 (or below $\nicefrac{1}{3}$), this can still be reported as “moderate” support in data for one hypothesis over the other. However, in doing so the researcher cannot any longer say that they are controlling the rate of false evidence at a prespecified level.

@@

\begin{tcolorbox}[title=\textbf{Practical recommendations 3},title filled]
{If the researcher has an informed idea on how the studied effect will look, they should consider using specification of H\textsubscript{1} other than the "default" zero-centered Cauchy employed in this tutorial. This could e.g., be a normal distribution centered around a value considered to be either a plausible group difference or is clinically meaningful. This recommendation also applies to the null-hypothesis; we encourage the researcher to consider redefining H\textsubscript{0} in order to test (what might be) a more appropriate hypothesis. An example of this could be to assess if an increase in patient BP\textsubscript{ND} gains more support in data compared to a decrease, rather than just testing "there is a difference" vs. "there is no difference".}
\end{tcolorbox}

In addition to the settings presented in this tutorial, there are several modifications that can be made for the BF test. For example, different decision thresholds can be applied for H~1~ and H~0~ depending on whether the researcher thinks that false positives are more important to avoid than false negatives, or vice versa. The researchers might also think, based on previous literature or bounds of a minimal effect size of interest, that a different specification of the alternative-hypothesis is more suitable to their research question (see "Practical Recommendations 3").

@@

It should by now be clear to the reader that a set of critical decisions needs to be made before applying sequential BF testing in a clinical PET study. For this reason, we recommend all researchers to pre-register their analysis before data collection starts [^fn10] [@VantVeer2016; @Poldrack2017; @Knudsen2020]. A pre-registration can be thought of as a safety net for the researcher. It helps guide the analysis and interpretation of data so that error rates are kept under control. It can also be shown to reviewers or readers to increase the credibility of the methods and findings. Deviations from a pre-registration is of course possible, and often warranted, but should be reported transparently in the article.

@@

If the reader is interested in learning more about sequential testing the following articles are a good start for a BF approach: @Rouder2014, @Schonbrodt2017 and @Schonbrodt2018; and for a NHST approach: @Schulz2005, @Proschan2006 and @Lakens2014.
